Fail Faster

### 2025-03-11 paper/paper.tex
continuing to explain

Values of \texttt{code} type are introduced by \emph{quotes}, written \texttt{.<$\ldots$>.}. Brackets
delay execution of a program until run time. For example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.
Note that this is not the same as \texttt{.< 6 >.}. Because brackets delay computation, the code is not \emph{executed}
until the next stage (run time).

Values of type \texttt{code} can be combined together using \emph{escape}, written \texttt{.\~{}($\ldots$)}.
Escape lets you take a value of type \texttt{code}, and ``splice'' it directly into a quote.
For example, this program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to \texttt{.<(1 * 5) + (1 * 5)>.}.

>>>>

Values of \texttt{code} type are introduced by \emph{quotes}, written
\texttt{.<$\ldots$>.}. Brackets delay execution of a program until run time. For
example, the program \texttt{.< 5 + 1 >.} has type \texttt{int code}.  Note that
this is not the same as \texttt{.< 6 >.}. Because brackets delay computation,
the code is not \emph{executed} until the next stage (run time).
Values of type \texttt{code} can be combined together using \emph{escape},
written \texttt{.\~{}($\ldots$)}.  Escape lets you take a value of type
\texttt{code}, and ``splice'' it directly into a quote.  For example, this
program \texttt{let x = .<1 * 5>. in .< .\~{}(x) + .\~{}(x) >.} evaluates to
\texttt{.<(1 * 5) + (1 * 5)>.}.  MetaOCaml ensures correct scoping and macro
hygiene, ensuring that variables are not shadowed when open terms are spliced.

The power of staging for optimization away abstraction overheads comes from
defining functions that accept and return \texttt{code} values.  A function
\texttt{f : 'a code -> 'b code} is a function that takes a program computing a
run-time \texttt{'a} and transforms it into a program computing a run-time
\texttt{'b}. In particular, because \texttt{f} itself runs at compile time, the
abstraction of using \texttt{f} is necessarily completely eliminated by
run-time.
A code-transforming function \texttt{'a code -> 'b code} can also be converted \emph{code for a function} --- a value of type
\texttt{('a -> 'b) code} --- with the following program:
\begin{lstlisting}
let eta (f : 'a code -> 'b code) : ('a -> 'b) code = .<fun x -> .\~(f .<x>.)>.
\end{lstlisting}

This program is known as ``the trick'' in the partial evaluation and multi-stage programming literature.
\cn{} 
% USE THIS: https://arxiv.org/pdf/2309.08207

\jwc{words}

As an example, consider
\begin{lstlisting}
let is_even x = .< .~x mod 2 == 0 >. in
let succ x = .< 1 + .~x >. in
eta (succ . is_even)
\end{lstlisting}
This is equal to \texttt{.< fun x -> (1 + x) mod 2 == 0 >.}: the functions have been fused together, and the overhead of calling them is not present at runtime.

This ``trick'' is how we use metaprogramming
